---
title: "CHL5233: Quarto in-class assignment"
subtitle: "Geographic coverage of the supplied country list"
author: "Milly Wu"
format:
  pdf:
    pdf-engine: xelatex
    documentclass: article
    fontsize: 10pt
    colorlinks: true
    geometry:
      - top=30mm
      - bottom=22mm
      - left=20mm
      - right=20mm
      - heightrounded
execute:
  echo: false
  warning: false
  message: false
  error: false
---

```{r}
#| label: setup
#| include: false
# include: false executes the setup while hiding its code and all output.
library(tidyverse)
library(knitr)
library(kableExtra)
options(dplyr.summarise.inform = FALSE)
region_colors <- c(
  Africa = "#0072B2", Americas = "#D55E00", Asia = "#009E73",
  Europe = "#CC79A7", Oceania = "#E69F00"
)
theme_set(theme_minimal(base_size = 11))
```

This report describes the countries in the supplied course files, matches them
using three-letter ISO codes, and summarizes their geographic distribution.
The analysis counts countries; it does not estimate conflict exposure or health
outcomes.

## Read the source data

```{r}
#| label: read-countries
#| echo: true
countries <- read.table(
  "../data/raw/countries.txt", header = TRUE,
  stringsAsFactors = FALSE, fileEncoding = "latin1"
)
dim(countries)
```

The country list contains **`r nrow(countries)` countries** and
**`r ncol(countries)` columns**, with `r n_distinct(countries$ISO)` distinct
ISO codes. The supplied list contains 186 countries, although the paper title
quoted in the assignment refers to 181 countries. This report uses the supplied
list without dropping countries to match that title.

```{r}
#| label: read-regions
#| echo: true
regions <- read.table(
  "../data/raw/regions.txt", header = TRUE,
  stringsAsFactors = FALSE, fileEncoding = "latin1"
)
dim(regions)
```

The region lookup contains **`r nrow(regions)` country or territory records** and
**`r ncol(regions)` columns**. The raw file includes imperfectly encoded country
names; ISO codes and geographic category labels are used for matching and
summaries. The original files are preserved unchanged.

## Create the analytical data set

```{r}
#| label: match-countries
#| echo: true
analysis_data <- regions |>
  semi_join(countries, by = c("country_code" = "ISO"))
unmatched <- countries |>
  anti_join(regions, by = c("ISO" = "country_code"))
```

```{r}
#| label: validate-matches
#| include: false
stopifnot(!anyDuplicated(countries$ISO),
          !anyDuplicated(regions$country_code),
          nrow(unmatched) == 0L,
          nrow(analysis_data) == nrow(countries),
          all(!is.na(analysis_data$region) & analysis_data$region != ""),
          all(!is.na(analysis_data$sub_region) & analysis_data$sub_region != ""))
```

All **`r nrow(analysis_data)` countries** match successfully; there are
**`r nrow(unmatched)` unmatched codes**. The analytical data cover
**`r n_distinct(analysis_data$region)` regions** and
**`r n_distinct(analysis_data$sub_region)` sub-regions**.

{{< pagebreak >}}

## Tabulate countries by region and sub-region

```{r}
#| label: count-regions
region_counts <- analysis_data |>
  group_by(region) |>
  summarise(Count = n(), .groups = "drop") |>
  arrange(desc(Count), region)

subregion_counts <- analysis_data |>
  group_by(region, sub_region) |>
  summarise(Count = n(), .groups = "drop") |>
  arrange(desc(Count), sub_region)

stopifnot(sum(region_counts$Count) == nrow(countries),
          sum(subregion_counts$Count) == nrow(countries))
```

Each country contributes once to each table. Both sets of counts sum to
**`r nrow(analysis_data)`**.

### Countries by region

```{r}
#| label: tbl-regions
#| tbl-cap: "Number of supplied study countries in each region."
kable(region_counts, col.names = c("Region", "Count"),
      booktabs = TRUE, align = c("l", "r")) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

### Countries by sub-region

```{r}
#| label: tbl-subregions
#| tbl-cap: "Number of supplied study countries in each sub-region."
subregion_counts |>
  select(sub_region, Count) |>
  kable(col.names = c("Sub-region", "Count"),
        booktabs = TRUE, align = c("l", "r")) |>
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

Africa contains the most included countries (**`r region_counts$Count[region_counts$region == "Africa"]`**),
while Oceania contains the fewest (**`r region_counts$Count[region_counts$region == "Oceania"]`**).
Sub-Saharan Africa has the largest sub-region count
(**`r subregion_counts$Count[subregion_counts$sub_region == "Sub-Saharan Africa"]`**).

{{< pagebreak >}}

## Visualize the number of countries by region

The chart below addresses the assignment's overall goal of comparing the number
of included countries across regions. Bars are ordered from the largest count
at the top to the smallest at the bottom.

```{r}
#| label: fig-regions
#| fig-cap: "Geographic coverage of the 186 countries in the supplied list."
#| fig-width: 7
#| fig-height: 3.4
#| fig-pos: H
#| echo: true
ggplot(analysis_data,
       aes(y = forcats::fct_infreq(region), fill = region)) +
  geom_bar(width = 0.65) +
  scale_y_discrete(limits = rev(region_counts$region)) +
  scale_fill_manual(values = region_colors, guide = "none") +
  scale_x_continuous(breaks = seq(0, 60, 10)) +
  labs(x = "Number of countries", y = "Region") +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

**Interpretation.** The country counts vary across regions. These bars describe
the composition of the supplied study list, not regional population size,
conflict prevalence, or the number of people affected by conflict.

{{< pagebreak >}}

## Visualize the number of countries by sub-region

Each bar represents one sub-region. Colors identify its parent region. Counts
run from largest at the top to smallest at the bottom; ties are ordered
alphabetically. `geom_bar()` counts the country rows directly.

```{r}
#| label: fig-subregions
#| fig-cap: "Countries by sub-region, ordered by count and colored by region."
#| fig-width: 7
#| fig-height: 5.5
#| fig-pos: H
subregion_order <- rev(subregion_counts$sub_region)
analysis_data <- analysis_data |>
  mutate(sub_region = factor(sub_region, levels = subregion_order))

ggplot(analysis_data, aes(y = sub_region, fill = region)) +
  geom_bar(width = 0.7) +
  geom_text(stat = "count", aes(label = after_stat(count)),
            hjust = -0.2, size = 3.2) +
  scale_fill_manual(values = region_colors) +
  scale_x_continuous(breaks = seq(0, 50, 10),
                     expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Number of countries", y = NULL, fill = "Region") +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())
```

**Interpretation.** Sub-Saharan Africa and Latin America and the Caribbean
contain `r subregion_counts$Count[subregion_counts$sub_region == "Sub-Saharan Africa"]`
and `r subregion_counts$Count[subregion_counts$sub_region == "Latin America and the Caribbean"]`
included countries, respectively. The chart uses the regional classification
provided in `regions.txt`; it does not redefine the sub-regions.

**Data source.** Course-provided `countries.txt` and `regions.txt`, associated
with the assignment on *Implications of armed conflict for maternal and child
health: A regression analysis of data from 181 countries for 2000-2019*.
